Characterizing dynamical transitions by statistical complexity measures based on ordinal pattern transition networks

### 2020-11-21 manuscript_Huang.tex
updated on 22 Nov.

\begin{figure*}
	\centering 
	\includegraphics[width=2\columnwidth]{logisticEntropy.pdf}
\caption{\small{Behavior of different statistical complexity measures for the Logistic map in dependence on the control parameter $r$. (a) bifurcation diagram, (b) the spectrum of the maximal Lyapunov exponent, (c) pattern frequency based (permutation) entropy $\mathcal{S}_O$ and (d) complexity $\mathcal{C}_O$; (e) pattern transition frequency based entropy $\mathcal{S}_W$ by the globally normalized weighting matrix $\mathbf{W}$ and (f) the corresponding complexity $\mathcal{C}_W$. (g, h) are entropy $\mathcal{S}_E$ and complexity $\mathcal{C}_E$ which are similar with (e, f) but based on node-wise out-link normalized weighting matrix $\mathbf{W}$. Several major periodic windows have been highlighted by grey background. Vertical dotted lines correspond to the cases as summarized in Tab. \ref{tableLog}. } \label{fig:bifurcation}}
\end{figure*}

@@

\section{Transitions in continuous system}
The previous results have been further verified in the chaotic R\"ossler system \cite{Roessler1976} while varying the control parameter, namely, $(\dot{x}, \dot{y}, \dot{z}) = (-y-z, x+0.2y, b+z(x-5.7))$, where $b$ is the control parameter. There are some challenging problems for constructing OPTNs for time series from continuous systems, for instance, proper choice of sampling frequency and embedding parameters, which must depend on the particular time scales of the system. To minimize the discussions on proper choices of further algorithmic parameters, we introduce Poincar\'e sections at $y=0$, $\dot{y}<0$ (for each value of $b$) and construct OPTNs from $N = 10000$ intersection points. The results have been shown in Fig. \ref{fig:bifurRossler}. 
\begin{figure*}
	\centering 
	\includegraphics[width=2\columnwidth]{rosslerEntropy.pdf}
\caption{\small{Similar with the caption of Fig. \ref{fig:bifurcation}. Behavior of different SCMs for the R\"ossler system varying the control parameter $b$. (a) bifurcation diagram ($x$-component on the Poinar\'e sections), (b) the maximal Lyapunov exponent $\lambda$, (c) pattern frequency based permutation entropy $\mathcal{S}_O$ and (d) complexity $\mathcal{C}_O$; (e) pattern transition frequency based entropy $\mathcal{S}_W$ by the globally normalized weighting matrix $\mathbf{W}$ and (f) the corresponding complexity $\mathcal{C}_W$. (g, h) are entropy $\mathcal{S}_E$ and complexity $\mathcal{C}_E$ which are similar with (e, f) but based on node-wise out-link normalized weighting matrix $\mathbf{W}$. Several major periodic windows have been highlighted by grey background. } \label{fig:bifurRossler}}
\end{figure*}
Rather similar results as Fig. \ref{fig:bifurcation} have been obtained in the chaotic R\"ossler system, identifying different dynamic transitions, where some periodic windows have been highlighted.

@@

\section{Conclusions} \label{sec:con}
In summary, we have proposed to include ordinal pattern transitions into the traditional ordinal pattern frequency analysis, generalizing statistical complexity measures for nonlinear time series analysis. The pattern transition properties of the corresponding time series networks provide novel insights for permutation entropy which relies on the static pattern frequency only. Therefore, this is another case showing the usefulness of ordinal pattern transition network for time series analysis \cite{ZouPR2018}. In addition, we suggest two slightly different ways to consider the effects of pattern transitions which are based on (i) a globally normalized weight transition matrix $\mathbf{W}$ and (ii) node-wise normalized transition matrix. Note that the node-wise based SCMs consider both individual pattern frequencies and and the transition frequencies among them. We show that these two new statistical complexity measures characterize the complexity properties of deterministic chaos successfully, far away from the lower right bottom of the complexity-entropy plane (Fig. \ref{fig:fBmNoise}). This helps to obtain deeper understandings of complexity analysis for time series of both deterministic chaos and stochastic systems, for instance, further numerical performance are required to show distinctions between different dynamics including various models of chaotic maps, long range correlated fractional Brownian motion and fractional Gaussian noise etc \cite{BorgesAMC2019,RossoPRE2007}. We have demonstrated the usefulness of SCMs by further results of time series for experimental fluid flows and ECG recordings, showing distinctive characteristics for different dynamics.

>>>>

\begin{figure*}
	\centering 
	\includegraphics[width=2\columnwidth]{logisticEntropy.pdf}
\caption{\small{Behavior of different statistical complexity measures for the Logistic map in dependence on the control parameter $r$. (a) bifurcation diagram, (b) the spectrum of the maximal Lyapunov exponent, (c) pattern frequency based (permutation) entropy $\mathcal{H}_O$ and (d) complexity $\mathcal{C}_O$; (e) pattern transition frequency based entropy $\mathcal{H}_W$ by the globally normalized weighting matrix $\mathbf{W}$ and (f) the corresponding complexity $\mathcal{C}_W$. (g, h) are entropy $\mathcal{H}_E$ and complexity $\mathcal{C}_E$ which are similar with (e, f) but based on node-wise out-link normalized weighting matrix $\mathbf{W}$. Several major periodic windows have been highlighted by grey background. Vertical dotted lines correspond to the cases as summarized in Tab. \ref{tableLog}. } \label{fig:bifurcation}}
\end{figure*}

@@

\section{Transitions in continuous system}
The previous results have been further verified in the chaotic R\"ossler system \cite{Roessler1976} while varying the control parameter, namely, $(\dot{x}, \dot{y}, \dot{z}) = (-y-z, x+0.2y, b+z(x-5.7))$, where $b$ is the control parameter. There are some challenging problems for constructing OPTNs for time series from continuous systems, for instance, proper choice of sampling frequency and embedding parameters, which must depend on the particular time scales of the system. To minimize the discussions on proper choices of further algorithmic parameters, we introduce Poincar\'e sections at $y=0$, $\dot{y}<0$ (for each value of $b$) and construct OPTNs from $N = 10000$ intersection points. The results have been shown in Fig. \ref{fig:bifurRossler}. 
\begin{figure*}
	\centering 
	\includegraphics[width=2\columnwidth]{rosslerEntropy.pdf}
\caption{\small{Similar with the caption of Fig. \ref{fig:bifurcation}. Behavior of different SCMs for the R\"ossler system varying the control parameter $b$. (a) bifurcation diagram ($x$-component on the Poinar\'e sections), (b) the maximal Lyapunov exponent $\lambda$, (c) pattern frequency based permutation entropy $\mathcal{H}_O$ and (d) complexity $\mathcal{C}_O$; (e) pattern transition frequency based entropy $\mathcal{H}_W$ by the globally normalized weighting matrix $\mathbf{W}$ and (f) the corresponding complexity $\mathcal{C}_W$. (g, h) are entropy $\mathcal{H}_E$ and complexity $\mathcal{C}_E$ which are similar with (e, f) but based on node-wise out-link normalized weighting matrix $\mathbf{W}$. Several major periodic windows have been highlighted by grey background. } \label{fig:bifurRossler}}
\end{figure*}
Rather similar results as Fig. \ref{fig:bifurcation} have been obtained in the chaotic R\"ossler system, identifying different dynamic transitions, where some periodic windows have been highlighted.

@@

\section{Conclusions} \label{sec:con}
In summary, we have proposed to include ordinal pattern transitions into the traditional ordinal pattern frequency analysis, generalizing statistical complexity measures for nonlinear time series analysis. The pattern transition properties of the corresponding time series networks provide novel insights for permutation entropy which relies on the static pattern frequency only. Therefore, this is another case showing the usefulness of ordinal pattern transition network for time series analysis \cite{ZouPR2018}. In addition, we suggest two slightly different ways to consider the effects of pattern transitions which are based on (i) a globally normalized weight transition matrix $\mathbf{W}$ and (ii) node-wise normalized transition matrix. Note that the node-wise based SCMs consider both individual pattern frequencies and the transition frequencies among them. We show that these two new statistical complexity measures characterize the complexity properties of deterministic chaos successfully, far away from the lower right bottom of the complexity-entropy plane (Fig. \ref{fig:fBmNoise}). This helps to obtain deeper understandings of complexity analysis for time series of both deterministic chaos and stochastic systems, for instance, further numerical studies are required to show distinctions of various dynamical systems of chaotic maps, long range correlated fractional Brownian motion and fractional Gaussian noise etc \cite{BorgesAMC2019,RossoPRE2007}. We have demonstrated the usefulness of SCMs by further results of time series for experimental fluid flows and ECG recordings, showing distinctive characteristics for different dynamics.
